---
title: "Getting and Cleaning Data - Course Project"
author: "Philip Mateescu"
date: "September 23, 2015"
output: html_document
---
Throughout this exercise we'll make use of two very useful libraries: *dplyr* and *tidyr*.
A lot of the work in this project is similar to the clean-up performed in the *swirl* exercise
on cleaning data using *tidyr*.

```{r}
library(dplyr)
library(tidyr)
```

## General Preparation

The general preparation applies to reading data that is common to both **test** and **train** datasets:
activity names and measurements performed/collected.

First lets load the activities as there's not much to clean up there.

```{r}
activities <- read.table('data/activity_labels.txt', sep = ' ', col.names = c('id', 'activityname'))
print(activities)
```

Features, aka what the measurements represent, have the same format, but the data will require
a bit of post-processing to clean it up.

```{r}
features <- read.table('data/features.txt', sep = ' ', col.names = c('id', 'feature'))
# print(features)
head(features)
```

Within the `r nrow(features)` rows, the `feature` column looks like a combination of three values: signal domain (time or frequency), signal (e.g. Body Acceleration, Gravity Acceleration), measurement type (mean, max, entropy), and direction, whether axis (X, Y, Z), or band (e.g 1-8, 9-16, etc).
We'll split this column into those respective columns in two passes: first we split into `signal`, `type`, and `direction` as these use a non alpha-numeric character;
then we extract the first character from `signal` into `domain` and the rest will remain in `signal` 

```{r, message=F, warning=F}
features <- features  %>%
    mutate(longname=feature) %>%
    separate(feature, into = c('signal', 'type', 'direction'), sep = '[^[:alnum:],]+') %>%
    separate(signal, into = c('domain', 'signal'), sep = 1)
# features are now tidy
head(features)
```

## Data-set specific preparation

Within each dataset, we have three files:

- **X_test.txt** - the measurements collected. There is one column for each of the rows in the **features** dataset.
- **subject_test.txt** - the id of the volunteer for which the data was collected. 
  There is one column only, and each row matches to one in **X_test.txt**.
- **y_test.txt** - the activity performed for each of the rows in the **X_test.txt** file.
  There is only one column and its values match against the **activities** dataset.
  
  
We'll start with the **test** dataset as it's smaller.

## Load *test* dataset

Read the subjects.
```{r}
test_subjects <- read.table('data/test/subject_test.txt', col.names = c('id'))
table(test_subjects)
```

Read the activities.
```{r}
test_activities <- read.table('data/test/y_test.txt', col.names = c('id'))
table(test_activities)
```

Now let's read the actual test data. The data seems to be fixed width, 
however, not all columns are of equals size: if columns that contain negative values
are 1 character larger than columns that have only positive values.
Calculating the precise widths for 561 columns could be tedious. 

Instead let's treat the file as if it's a space-separated value list.
Initial attempts to read the file as is, shows us running into issues as some 
columns are separated by one space, others by two.

So we'll pre-process the file, outside of this script, to replace all multiple spaces
with single spaces, then proceed to read it as a space-separated table.

```{r}
test_data <- read.table('data/test/10X_test.txt', sep = ' ')
```

It's a large object: `r format(object.size(test_data), units = 'MB')`MB.

Since we didn't provide any column names for our data, we're being provided the standard
*V+number*:

```{r}
str(colnames(test_data))
```

## Cleanup *test* dataset

```{r}
# temporary
test_subjects <- sample(unique(test_subjects$id), nrow(test_data), replace=T)
```
```{r}
test_activities <- sample(unique(test_activities$id), nrow(test_data), replace=T)
```

First let's insert one column for each *subject*, *activity*.

```{r}
test_data <- mutate(test_data, subjectid=test_subjects, activityid=test_activities)
```

Now let's gather all the measurements into a column
```{r}
test_data <- test_data %>%
             gather(key=featureid, value=value, -subjectid, -activityid) %>%
             mutate(featureid=extract_numeric(featureid)) 
summary(test_data)
```

Let's write all this in a function since we'll be using it later:

```{r}
read_signals <- function(dataset_name, debug = F) {
    folder <- paste('data', dataset_name, sep = '/')
    subjects_file <- paste(folder, '/subject_', dataset_name, '.txt', sep = '')
    activities_file <- paste(folder, '/y_', dataset_name, '.txt', sep = '')
    data_file_prefix <- '/X_'
    if (debug) {
        data_file_prefix <- '/10X_'
    }
    data_file <- paste(folder, data_file_prefix, dataset_name, '.txt', sep = '')
    subjects <- read.table(subjects_file, col.names = c('id'))
    activities <- read.table(activities_file, col.names = c('id'))
    data <- read.table(data_file, sep = ' ')
    if (debug) {
        subids <- sample(unique(subjects$id), nrow(data), replace=T)
        actids <- sample(unique(activities$id), nrow(data), replace=T)
    } else {
        subids <- subjects$id
        actids <- activities$id
    }
    # First let's insert one column for each *subject*, *activity*.
    data <- data %>%
         mutate(subjectid=subids, activityid=actids) %>%
         # now let's gather all the measurements into a column
         gather(key=featureid, value=value, -subjectid, -activityid) %>%
         mutate(featureid=extract_numeric(featureid)) 
    
    return(data)
}
```

Now let's do the same with the **train** dataset using the function we created.

## Load and clean *train* dataset

Same assumptions as with the test dataset: we cleaned up the **X_train.txt** to remove double spaces.

```{r}
test_data <- read_signals('test', debug = T)
train_data <- read_signals('train', debug = T)
```

The **train** dataset is larger: `r format(object.size(test_data), units = 'MB')`MB.


# Assignment Tasks

## Task 1 - merge datasets

Merge the training and the test sets to create one data set.

```{r}
data <- bind_rows(test_data, train_data)
```

## Task 2 - extract mean and deviation

Extract only the measurements on the mean and standard deviation for each measurement

```{r}
# print(features$longname)
ids_of_mean_and_average <- grep('(mean\\(|avg\\())', features$longname)
mean_and_dev <- filter(data, featureid %in% ids_of_mean_and_average)
```
We have **`r format(nrow(ids_of_mean_and_average))` measurements**.

## Task 3 - descriptive activity names

Use descriptive activity names to name the activities in the data set.

```{r}
mean_and_dev <- merge(mean_and_dev, activities, by.x='activityid', by.y='id')
mean_and_dev <- merge(mean_and_dev, features, by.='featureid', by.y='id')
```

## Task 4 - use descriptive variable names

Appropriately label the data set with descriptive variable names.

```{r}
# the columns have already been named, let's name the measurements more explicitly
# let's trim the feature name a bit
mean_and_dev$longname <- gsub('\\(\\)', '', mean_and_dev$longname)
mean_and_dev$longname <- gsub('-', '.', mean_and_dev$longname)
mean_and_dev$activityname <- gsub('(\\w)(\\w+)', '\\U\\1\\L\\2', mean_and_dev$activityname, perl=TRUE)
mean_and_dev$activityname <- gsub('_', ' ', mean_and_dev$activityname)
head(mean_and_dev)
```


## Task 5 - extract average for each activity and subject

From the data set in step 4, create a second, independent tidy data set with the average of each variable for each activity and each subject.

```{r}
tidy <- mean_and_dev %>%
     group_by(subjectid, activityname, meansurementtype=longname) %>%
     summarize(averagemeasurementvalue=mean(value))
print(tidy)
```
    




